Modelling state-switching functional data with hidden Markov models

### 2020-11-24 Report/Sections/discussion.tex
Read through main text

Current Functional Data Analysis literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}. Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
We deal with temporal dependence between curves with distinct ``types" by using either an HMM or a CarHMM to model the curve sequence. We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model. Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}. We also incorporate a moving-window transformation into the fine-scale model to deal with intricate dependence structures on short time scales. Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales. Provided the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.

We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast British Columbia, Canada. This gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of the Northern Resident killer whales \citep{Noren:2011}. This model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}. In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.

>>>>

Current Functional Data Analysis literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}. Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
We deal with temporal dependence between curves by using either an HMM or a CarHMM to model the curve sequence. We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model. Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}. We also incorporate a moving-window transformation into the fine-scale model to deal with intricate dependence structures on short time scales. Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales. Provided the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.

We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast British Columbia, Canada. This gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of northern resident killer whales \citep{Noren:2011}. This model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}. In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.

### 2020-11-24 Report/Sections/simulation.tex
Read through main text

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

The two simulated dive types differ in that dives of type one are much shorter on average (26 seconds) than dives of type two (105 seconds). The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data. All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. The average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9. All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to underestimate the empirical standard errors. This is especially true for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error could be up to twice as large as the estimated standard error. See Tables 5 through 9 of the supplementary material for detailed results.

The HHMM-DFT performs similarly to the CarHHMM-DFT in most respects. Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is approximately $5 \%$ worse. Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material). These results suggest that the estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored. This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.

@@

Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model cannot estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 24 of the supplementary material).

Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display one simulated data set as well as the decoded dive types and subdive states associated with each model. The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2 (higher activity). The CarHHMM is the least accurate model when predicting subdive state. The CarHHMM-DFT and HHMM-DFT output similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the second dive. The CarHMM-DFT does not estimate dive type.

>>>>

We generate 500 independent data sets using the CarHHMM-DFT as a generative model. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$. Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
    %
    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
    %
    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

The two simulated dive types differ in that dives of type one are much shorter on average (26 seconds) than dives of type two (105 seconds). The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$ since both are much higher for subdive state 2 than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data. All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}). If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives. Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows. We find both the average dive decoding accuracy and the average subdive decoding accuracy using the \textit{forward-backward} algorithm \citep{Zucchini:2016}.

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. Its average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9. All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases. In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors. This is especially true for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error. See Tables 5 through 9 of the supplementary material for detailed results.

The HHMM-DFT performs similarly to the CarHHMM-DFT in most respects. Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is worse by approximately 5 percentage points. Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$. In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material). These results suggest that the estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored. This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.

@@

Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model cannot estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 24 of the supplementary material).

Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display one simulated data set as well as the decoded dive types and subdive states associated with each model. The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2. As expected, the CarHHMM is the least accurate model when predicting subdive state. The CarHHMM-DFT and HHMM-DFT yield similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the second dive. The CarHMM-DFT does not estimate dive type.
